### 2018-07-05 paper/4-build.tex
Revise S4.2.1: A dirty bit

% An alternative approach to suspending a task is to abort
% it and restart it again later, at the cost of doing additional work.

@@

Suppose the scheduler decides that that a key \hs{k} should be
brought up to date.  The next question is: does any work need to be done,
or is it already up to date?   Or, in a cloud-build system, do we have
a cached copy of the value we need?
These questions can be addressed in one of four
fundamental ways, with a number of tweaks and variations within them.

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two states is marked dirty.
% ; and by marking additional things dirty/clean the build system can track
% what needs to rebuild.
If a key and all its transitive dependencies are clean, the key does not need
to be rebuild.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
\simon{Not always (see make) and only possible if we have static dependencies
  or have recorded dynamic dependencies.}
When rebuilding, if a clean cell only depends on clean cells it is skipped,
otherwise it is rebuilt. \simon{But if all the dependencies of a dirty cell
  are marked dirty, we rebuild only dirty cells not clean ones, contradicting this statement.}

% AM: I didn't understand this bit, so commented it out.
% The only wrinkle in this scheme is that \Excel supports
% monadic tasks, and does not separately record the rebuilt, so it has
% to approximate in this case.

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under the control of \Make; rather it is
existing file-system information that has been repurposed. In particular,
modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software. \simon{Also transitive dependencies
  of dirty files are not made dirty.}

When using a dirty bit, it is necessary to check all the dependencies of a key.
For applicative build systems that list is easy to obtain, but for monadic
build systems there is no general way to get all dependencies. Instead \Excel
computes a \emph{static approximation} of the dependencies. For applicative
tasks that approximation is correct. For functions such as \cmd{IF} it marks the
cell dirty if \emph{any} potential dependency has changed, even on the untaken
\cmd{IF} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
be guessed, it conservatively assumes the dependencies have always changed.

With a dirty bit it is simple to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
bit after a computation that did not change the value. \simon{And if we'd recursively
made the transitive dependencies dirty, we could not do early cutoff.}
\Excel could use this
approach, but does not. \simon{It'd be tricky becase of setting dirtiness recursively.}
In contrast, \Make cannot implement early cutoff nicely
-- to do so it would have to mark the key clean (so it would not rebuild in the next
run) and at the same time not mark the keys that depend on it dirty -- an
impossible task with only the ability to update to the latest modification time.
\Make can approximate early cutoff by not modifying the result file, and not
marking it clean, but then it will rerun in every subsequent build.

>>>>

@@

Suppose the scheduler decides that that a key should be brought up to date. The
next question is: does any work need to be done, or is it already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?
These questions can be addressed in one of four fundamental ways, with a number
of tweaks and variations within them.

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuild.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
To mark dependent cells dirty, it is necessary to statically know task
dependencies. This is possible only for applicative tasks (see \S\ref{sec-deps}).
For monadic tasks, \Excel computes a \emph{static approximation} of task
dependencies: for functions such as \cmd{IF}, the approximation includes all
cells that are mentioned in the formula (the \cmd{IF}'s condition and both
branches), whereas functions such as \cmd{INDIRECT} whose dependencies cannot be
guessed, are conservatively assumed to depend on all cells and are therefore
always marked dirty.

\Make uses file modification times, and compares files to their dependencies,
which can be thought of as a dirty bit which is set when a file is older than
its dependencies. The interesting property of this dirty bit is that it is not
under the control of \Make; rather it is existing file-system information that
has been repurposed. Modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the files depending on it (but not
recursively). Note that \Make requires that file timestamps only go forward in
time, which can be violated by backup software.

With a dirty bit it is simple to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
bit after a computation that did not change the value and make sure that cells
that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
because the dependent cells have already been recursively marked dirty. For
\Make, it is impossible to mark a file clean and at the same time not mark the
keys that depend on it dirty. \Make can approximate early cutoff by not
modifying the result file, and not marking it clean, but then it will be rebuilt
in every subsequent build.
